
---
title: "Data Mining assignment Report"
subtitle: "Income Prediction (DT, NB, LR, KNN, ANN) | Iris Clustering | Outlier Detection"
author: "Maganda Jonathan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
# Load all required packages (merged from both analyses)
library(tidyverse)      # data manipulation & plotting
library(caret)          # train/test split, confusion matrix, preprocessing
library(rpart)          # decision trees
library(rpart.plot)     # plot trees
library(e1071)          # naive Bayes
library(pROC)           # ROC curves
library(knitr)          # nice tables
library(kableExtra)     # table styling
library(FNN)            # fast KNN (KD‑tree)
library(nnet)           # neural network (ANN)
library(fastDummies)    # one‑hot encoding for ANN
library(NeuralNetTools) # plot ANN
library(cluster)        # silhouette for clustering
library(dendextend)     # dendrogram colouring
library(isotree)        # isolation forest
library(dbscan)         # LOF

# Set global seed for reproducibility
set.seed(123)

# Global knitr options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

# 1. Introduction

This report covers:

- **Income prediction** from the Adult Census dataset using five classification models:
  Decision Tree, Naive Bayes, Logistic Regression, K‑Nearest Neighbors (KNN), and an Artificial Neural Network (ANN).
- **Clustering analysis** on the Iris dataset (k‑means and hierarchical).
- **Outlier detection** on the Adult dataset using four methods (IQR, Z‑score, Isolation Forest, LOF) with a consensus approach.
- A simple **One Rule (1R)** demonstration using a weather dataset.

All code is optimized for efficiency and readability.

# 2. Load and Prepare the Adult Dataset

```{r load-adult}
# Adjust the path to your local file
adult <- read.csv("C:/Users/MAGADA/Desktop/adult.csv", header = FALSE, na.strings = " ?")

# Assign column names
colnames(adult) <- c(
  "age", "workclass", "fnlwgt", "education", "education_num",
  "marital_status", "occupation", "relationship", "race",
  "sex", "capital_gain", "capital_loss", "hours_per_week",
  "native_country", "income"
)

# Remove leading/trailing spaces
adult <- adult %>% mutate(across(where(is.character), trimws))

# Initial inspection
glimpse(adult)
```

## 2.1 Data Quality Assessment

```{r data-quality}
dim(adult)
summary(adult)
colSums(is.na(adult))
```

## 2.2 Data Cleaning

Impute missing categorical values with the mode and numeric values with the mean.

```{r missing-values}
# Function to compute mode
getmode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

# Impute categorical columns (mode)
adult <- adult %>%
  mutate(across(c(workclass, occupation, native_country),
                ~ replace_na(., getmode(.[!is.na(.)]))))

# Impute numeric columns (mean)
adult <- adult %>%
  mutate(across(c(age, fnlwgt, education_num, capital_gain, capital_loss, hours_per_week),
                ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))
```

## 2.3 Factor Conversion

```{r factor-conversion}
adult <- adult %>%
  mutate(across(where(is.character), as.factor))

# Rename income levels for clarity
levels(adult$income) <- c("below_50K", "above_50K")
```

## 2.4 Outlier Identification (Initial Exploration)

```{r outliers}
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR <- Q3 - Q1
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  sum(x < lower | x > upper)
}

vars <- c("age", "hours_per_week", "capital_gain", "capital_loss")
map_dbl(adult[vars], detect_outliers)

# Boxplots
adult %>%
  select(all_of(vars)) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(y = value)) +
  geom_boxplot() +
  facet_wrap(~ name, scales = "free") +
  theme_minimal()
```

## 2.5 Exploratory Data Analysis

```{r eda-plots}
# Income distribution
ggplot(adult, aes(income, fill = income)) +
  geom_bar() +
  theme_minimal()

# Age density by income
ggplot(adult, aes(age, fill = income)) +
  geom_density(alpha = 0.5) +
  theme_minimal()

# Hours per week histogram
ggplot(adult, aes(hours_per_week, fill = income)) +
  geom_histogram(bins = 42, alpha = 0.6) +
  theme_minimal()

# Correlation heatmap of numeric variables
numeric_vars <- adult %>% select(age, education_num, capital_gain, capital_loss, hours_per_week)
cor_matrix <- cor(numeric_vars)

cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "cor") %>%
  ggplot(aes(var1, var2, fill = cor)) +
  geom_tile() +
  scale_fill_gradient2(low = "green", high = "yellow", mid = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2.6 Train‑Test Split

```{r train-test-split}
set.seed(123)
train_index <- createDataPartition(adult$income, p = 0.7, list = FALSE)

train <- adult[train_index, ]
test  <- adult[-train_index, ]
```

# 3. Income Prediction Models

## 3.1 Decision Tree 

```{r decision-tree}
tree_model <- rpart(income ~ ., data = train, method = "class")
rpart.plot(tree_model)
```

### Tree Pruning

```{r prune-tree}
printcp(tree_model)
best_cp <- tree_model$cptable[which.min(tree_model$cptable[, "xerror"]), "CP"]
pruned_model <- prune(tree_model, cp = best_cp)
rpart.plot(pruned_model)
```

### Evaluation

```{r dt-evaluation}
dt_pred <- predict(pruned_model, test, type = "class")
dt_prob <- predict(pruned_model, test, type = "prob")[, "above_50K"]

cm_dt <- confusionMatrix(dt_pred, test$income)

dt_accuracy    <- cm_dt$overall["Accuracy"]
dt_precision   <- cm_dt$byClass["Pos Pred Value"]
dt_recall      <- cm_dt$byClass["Sensitivity"]
dt_specificity <- cm_dt$byClass["Specificity"]
dt_f1          <- 2 * (dt_precision * dt_recall) / (dt_precision + dt_recall)
dt_auc         <- auc(roc(test$income, dt_prob, levels = rev(levels(test$income))))

cm_dt
```

## 3.2 Naive Bayes

```{r naive-bayes}
nb_model <- naiveBayes(income ~ ., data = train)

nb_pred <- predict(nb_model, test)
nb_prob <- predict(nb_model, test, type = "raw")[, "above_50K"]

cm_nb <- confusionMatrix(nb_pred, test$income)

nb_accuracy    <- cm_nb$overall["Accuracy"]
nb_precision   <- cm_nb$byClass["Pos Pred Value"]
nb_recall      <- cm_nb$byClass["Sensitivity"]
nb_specificity <- cm_nb$byClass["Specificity"]
nb_f1          <- 2 * (nb_precision * nb_recall) / (nb_precision + nb_recall)
nb_auc         <- auc(roc(test$income, nb_prob, levels = rev(levels(test$income))))

cm_nb
```

## 3.3 Logistic Regression

```{r logistic}
log_model <- glm(income ~ ., data = train, family = binomial)

log_prob <- predict(log_model, test, type = "response")
log_pred <- factor(ifelse(log_prob > 0.5, "above_50K", "below_50K"),
                   levels = levels(test$income))

cm_log <- confusionMatrix(log_pred, test$income)

log_accuracy    <- cm_log$overall["Accuracy"]
log_precision   <- cm_log$byClass["Pos Pred Value"]
log_recall      <- cm_log$byClass["Sensitivity"]
log_specificity <- cm_log$byClass["Specificity"]
log_f1          <- 2 * (log_precision * log_recall) / (log_precision + log_recall)
log_auc         <- auc(roc(test$income, log_prob, levels = rev(levels(test$income))))

cm_log
```

## 3.4 K‑Nearest Neighbors (Optimised with KD‑Tree)

KNN requires scaling and dummy coding.

```{r knn-prep}
# Create dummy variables for all predictors
dummies <- dummyVars(income ~ ., data = train)
train_dummy <- predict(dummies, newdata = train)
test_dummy  <- predict(dummies, newdata = test)

# Remove near‑zero variance predictors
nzv <- nearZeroVar(train_dummy)
if (length(nzv) > 0) {
  train_dummy <- train_dummy[, -nzv]
  test_dummy  <- test_dummy[, -nzv]
}

# Scale and center
pre_proc <- preProcess(train_dummy, method = c("center", "scale"))
train_scaled <- predict(pre_proc, train_dummy)
test_scaled  <- predict(pre_proc, test_dummy)

train_y <- train$income
test_y  <- test$income
```

```{r knn-model}
k_value <- 7  # can be tuned

# KD‑tree search
knn_fast <- get.knnx(data = train_scaled, query = test_scaled,
                     k = k_value, algorithm = "kd_tree")

# Majority vote
neighbor_labels <- matrix(train_y[knn_fast$nn.index], ncol = k_value)
knn_pred <- apply(neighbor_labels, 1, function(x) {
  names(sort(table(x), decreasing = TRUE))[1]
})
knn_pred <- factor(knn_pred, levels = levels(test_y))

cm_knn <- confusionMatrix(knn_pred, test_y)

knn_accuracy    <- cm_knn$overall["Accuracy"]
knn_precision   <- cm_knn$byClass["Pos Pred Value"]
knn_recall      <- cm_knn$byClass["Sensitivity"]
knn_specificity <- cm_knn$byClass["Specificity"]
knn_f1          <- 2 * (knn_precision * knn_recall) / (knn_precision + knn_recall)

cm_knn
```

## 3.5 Artificial Neural Network (ANN)

We train a single‑hidden‑layer neural network using the `nnet` package.  
**Note:** The ANN uses the same training/test split as above but requires its own preprocessing (one‑hot encoding and normalisation) because the original `train` data still contains factors.

```{r ann-prep}
# Use a fresh copy for ANN preprocessing
ann_train <- train
ann_test  <- test

# One‑hot encode categorical predictors (exclude income)
cat_cols <- c("workclass", "marital_status", "occupation", "relationship", "race", "sex", "native_country")
ann_train_encoded <- dummy_cols(ann_train, select_columns = cat_cols,
                                remove_first_dummy = TRUE, remove_selected_columns = TRUE)
ann_test_encoded  <- dummy_cols(ann_test,  select_columns = cat_cols,
                                remove_first_dummy = TRUE, remove_selected_columns = TRUE)

# Ensure column names are syntactically valid
colnames(ann_train_encoded) <- make.names(colnames(ann_train_encoded), unique = TRUE)
colnames(ann_test_encoded)  <- make.names(colnames(ann_test_encoded),  unique = TRUE)

# Normalise numeric predictors
normalize <- function(x) (x - min(x)) / (max(x) - min(x))
num_cols <- c("age", "education_num", "capital_gain", "capital_loss", "hours_per_week", "fnlwgt")
ann_train_encoded[num_cols] <- lapply(ann_train_encoded[num_cols], normalize)
ann_test_encoded[num_cols]  <- lapply(ann_test_encoded[num_cols],  normalize)

# Target as factor 0/1 (nnet formula interface expects factor)
ann_train_encoded$income <- factor(ifelse(ann_train_encoded$income == "above_50K", "1", "0"),
                                   levels = c("0", "1"))
ann_test_encoded$income  <- factor(ifelse(ann_test_encoded$income  == "above_50K", "1", "0"),
                                   levels = c("0", "1"))
```

```{r train-ann}
set.seed(42)
ann_model <- nnet(
  income ~ .,
  data = ann_train_encoded,
  size = 10,
  decay = 0.001,
  maxit = 300,
  trace = FALSE
)

cat("ANN architecture:", ann_model$n[1], "→", ann_model$n[2], "→", ann_model$n[3], "\n")
cat("Final cross‑entropy loss:", round(ann_model$value, 4), "\n")
```

### Network Visualisation

```{r network-plot, fig.height=8, fig.width=10}
plotnet(ann_model, alpha.val = 0.6,
        circle_col = list("lightblue", "lightgreen"),
        pos_col = "darkblue", neg_col = "darkred")
title("ANN Architecture: Input → 10 Hidden → 1 Output (logistic)", line = 2.5)
```

### ANN Evaluation

```{r ann-evaluation}
ann_prob <- predict(ann_model, ann_test_encoded, type = "raw")
ann_pred <- ifelse(ann_prob > 0.5, "1", "0")
ann_actual <- ann_test_encoded$income

cm_ann <- confusionMatrix(factor(ann_pred, levels = c("0", "1")), ann_actual, positive = "1")

ann_accuracy    <- cm_ann$overall["Accuracy"]
ann_precision   <- cm_ann$byClass["Pos Pred Value"]
ann_recall      <- cm_ann$byClass["Sensitivity"]
ann_specificity <- cm_ann$byClass["Specificity"]
ann_f1          <- 2 * (ann_precision * ann_recall) / (ann_precision + ann_recall)
ann_auc         <- auc(roc(ann_actual, ann_prob, levels = rev(levels(ann_actual))))

cm_ann
```

## 3.6 ROC Curve Comparison (All Models)

```{r roc}
roc_dt  <- roc(test$income, dt_prob,  levels = rev(levels(test$income)))
roc_nb  <- roc(test$income, nb_prob,  levels = rev(levels(test$income)))
roc_log <- roc(test$income, log_prob, levels = rev(levels(test$income)))
roc_ann <- roc(ann_actual, ann_prob, levels = rev(levels(ann_actual)))

plot(roc_dt, col = "grey", main = "ROC Curve Comparison")
lines(roc_nb, col = "yellow")
lines(roc_log, col = "green")
lines(roc_ann, col = "red")
legend("bottomright",
       legend = c("Decision Tree", "Naive Bayes", "Logistic Regression", "ANN"),
       col = c("grey", "yellow", "green", "red"), lwd = 2)
```

# 4. Clustering Analysis – Iris Dataset

```{r iris-load}
data(iris)
iris_features <- iris[, 1:4]
iris_scaled <- as.data.frame(scale(iris_features))
```

## 4.1 Optimal Number of Clusters

```{r optimal-k}
set.seed(123)
wss_values <- sapply(1:10, function(k) {
  kmeans(iris_scaled, centers = k, nstart = 25)$tot.withinss
})

wss_df <- data.frame(k = 1:10, wss = wss_values)
ggplot(wss_df, aes(x = k, y = wss)) +
  geom_line(color = "#3598DB", linewidth = 1) +
  geom_point(color = "#E73C3C", size = 3) +
  geom_vline(xintercept = 3, linetype = "dashed", color = "gray42") +
  labs(title = "Elbow Method – Optimal k = 3", x = "k", y = "Within‑cluster sum of squares") +
  theme_minimal()
```

## 4.2 K‑Means Clustering

```{r kmeans}
set.seed(42)
kmeans_model <- kmeans(iris_scaled, centers = 3, nstart = 25)
cat("Cluster sizes:", kmeans_model$size, "\n")
cat("BSS/TSS ratio:", round(kmeans_model$betweenss / kmeans_model$totss * 100, 2), "%\n")
```

## 4.3 Hierarchical Clustering

```{r hierarchical}
dist_matrix <- dist(iris_scaled, method = "euclidean")
hc_ward <- hclust(dist_matrix, method = "ward.D2")
hc_clusters <- cutree(hc_ward, k = 3)

dend <- as.dendrogram(hc_ward)
dend <- color_branches(dend, k = 3, col = c("#E73C3C", "#3488DB", "#2EdC71"))
plot(dend, main = "Dendrogram – Ward's Linkage", leaflab = "none")
abline(h = 5.5, lty = 2, col = "gray42")
```

## 4.4 Cluster Validation

```{r validation}
table(KMeans = kmeans_model$cluster, Species = iris$Species)
table(Hierarchical = hc_clusters, Species = iris$Species)

sil_km <- silhouette(kmeans_model$cluster, dist_matrix)
sil_hc <- silhouette(hc_clusters, dist_matrix)
cat("K‑Means Avg Silhouette :", round(mean(sil_km[, 3]), 4), "\n")
cat("Hierarchical Avg Silhouette:", round(mean(sil_hc[, 3]), 4), "\n")
```

## 4.5 PCA Visualisation

```{r pca-viz}
pca_result <- prcomp(iris_scaled)
pca_coords <- as.data.frame(pca_result$x[, 1:2])
pca_coords$cluster <- as.factor(kmeans_model$cluster)
pca_coords$species <- iris$Species

ggplot(pca_coords, aes(x = PC1, y = PC2, color = cluster, shape = species)) +
  geom_point(size = 3, alpha = 0.85) +
  scale_color_manual(values = c("#E73C3C", "#3488DB", "#2EdC71")) +
  stat_ellipse(aes(group = cluster), linetype = "dashed") +
  labs(title = "K‑Means Clusters – PCA Projection",
       x = paste0("PC1 (", round(summary(pca_result)$importance[2, 1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_result)$importance[2, 2] * 100, 1), "%)")) +
  theme_minimal()
```

# 5. Outlier Detection on the Adult Dataset

We use four methods on the original numeric features.

```{r outlier-data}
# Select numeric columns from the original (cleaned) adult data
adult_num <- adult[, sapply(adult, is.numeric), drop = FALSE]
```

## 5.1 IQR

```{r iqr}
iqr_outliers <- function(x, multiplier = 1.5) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_val <- Q3 - Q1
  lower <- Q1 - multiplier * IQR_val
  upper <- Q3 + multiplier * IQR_val
  return(list(flag = x < lower | x > upper, n_out = sum(x < lower | x > upper)))
}

iqr_results <- lapply(adult_num, iqr_outliers)
iqr_summary <- data.frame(
  Feature = names(adult_num),
  N_Outliers = sapply(iqr_results, function(x) x$n_out),
  Pct = round(sapply(iqr_results, function(x) x$n_out / nrow(adult_num) * 100), 2)
)
print(iqr_summary)
```

## 5.2 Z‑Score

```{r zscore}
z_scores <- as.data.frame(scale(adult_num))
z_flag <- abs(z_scores) > 3
z_summary <- data.frame(
  Feature = names(adult_num),
  N_Outliers = colSums(z_flag),
  Pct = round(colMeans(z_flag) * 100, 2),
  Max_Z = round(apply(abs(z_scores), 2, max, na.rm = TRUE), 2)
)
print(z_summary)
```

## 5.3 Isolation Forest

```{r isolation}
set.seed(42)
adult_num_clean <- na.omit(adult_num)

iso_model <- isolation.forest(adult_num_clean, ntrees = 100, sample_size = min(256, nrow(adult_num_clean)))
iso_scores <- predict(iso_model, adult_num_clean, type = "score")
iso_flag <- iso_scores > 0.55

cat("Isolation Forest anomalies:", sum(iso_flag), "/", nrow(adult_num_clean),
    sprintf("(%.2f%%)\n", mean(iso_flag) * 100))

iso_df <- data.frame(score = iso_scores, flag = iso_flag)
ggplot(iso_df, aes(x = score, fill = flag)) +
  geom_histogram(bins = 60, color = "white", alpha = 0.85) +
  geom_vline(xintercept = 0.55, color = "#E73C3C", linetype = "dashed") +
  scale_fill_manual(values = c("#3488DB", "#E73C3C"), labels = c("Normal", "Anomaly")) +
  labs(title = "Isolation Forest Anomaly Scores", x = "Anomaly Score", y = "Count") +
  theme_minimal()
```

## 5.4 Local Outlier Factor (LOF)

```{r lof}
set.seed(42)
adult_scaled_mat <- scale(adult_num_clean)
lof_scores <- lof(adult_scaled_mat, minPts = min(20, nrow(adult_scaled_mat) - 1))
lof_flag <- lof_scores > 2

cat("LOF anomalies:", sum(lof_flag), "/", nrow(adult_num_clean),
    sprintf("(%.2f%%)\n", mean(lof_flag) * 100))

lof_df <- data.frame(lof = pmin(lof_scores, 10, na.rm = TRUE), flag = lof_flag)
ggplot(lof_df, aes(x = lof, fill = flag)) +
  geom_histogram(bins = 60, color = "white", alpha = 0.85) +
  geom_vline(xintercept = 2, color = "#E74C3C", linetype = "dashed") +
  scale_fill_manual(values = c("#3498DB", "#E74C3C"), labels = c("Normal", "Anomaly")) +
  labs(title = "LOF Scores (capped at 10)", x = "LOF Score", y = "Count") +
  theme_minimal()
```

## 5.5 Consensus Outliers

```{r consensus}
n_original <- nrow(adult)
iqr_flag_all <- rowSums(sapply(iqr_results, function(x) x$flag)) > 0
z_flag_all <- rowSums(z_flag, na.rm = TRUE) > 0

clean_idx <- which(complete.cases(adult_num))
iso_flag_all <- rep(FALSE, n_original)
lof_flag_all <- rep(FALSE, n_original)
iso_flag_all[clean_idx] <- iso_flag
lof_flag_all[clean_idx] <- lof_flag

adult$iqr_flag <- iqr_flag_all
adult$z_flag   <- z_flag_all
adult$iso_flag <- iso_flag_all
adult$lof_flag <- lof_flag_all

adult$n_methods <- (adult$iqr_flag + adult$z_flag + adult$iso_flag + adult$lof_flag)
adult$consensus_outlier <- adult$n_methods >= 2

cat("Flagged by 2+ methods (consensus):", sum(adult$consensus_outlier),
    sprintf("(%.2f%%)\n", mean(adult$consensus_outlier) * 100))

ggplot(adult, aes(x = factor(n_methods))) +
  geom_bar(fill = "#3498DB", color = "white") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "Outlier Detection Consensus",
       x = "Number of Methods Flagging", y = "Count") +
  theme_minimal()
```

# 6. One Rule (1R) Example Using Weather Dataset

```{r oner-weather}
# Adjust path to your weather.csv
weather <- read.csv("C:/Users/MAGADA/Desktop/weather.csv")
kable(weather)

oneR_error <- function(data, attribute, target = "play") {
  tab <- table(data[[attribute]], data[[target]])
  results <- data.frame(
    Value = rownames(tab),
    Yes = tab[, "yes"],
    No = tab[, "no"]
  )
  results$Majority_Class <- ifelse(results$Yes >= results$No, "yes", "no")
  results$Errors <- ifelse(results$Yes >= results$No, results$No, results$Yes)
  total_error <- sum(results$Errors)
  list(table = results, error = total_error)
}

outlook_results     <- oneR_error(weather, "outlook")
temperature_results <- oneR_error(weather, "temperature")
humidity_results    <- oneR_error(weather, "humidity")
windy_results       <- oneR_error(weather, "windy")

comparison <- data.frame(
  Attribute = c("Outlook", "Temperature", "Humidity", "Windy"),
  Errors = c(outlook_results$error, temperature_results$error,
             humidity_results$error, windy_results$error)
)
comparison$Accuracy <- (nrow(weather) - comparison$Errors) / nrow(weather) * 100

kable(comparison, caption = "1R Performance by Attribute")

best_attribute <- comparison$Attribute[which.min(comparison$Errors)]
cat("Best attribute for 1R rule:", best_attribute, "\n")
```

# 7. Model Comparison Table

```{r results-table}
results <- data.frame(
  Model       = c("Decision Tree", "Naive Bayes", "Logistic Regression", "KNN", "ANN"),
  Accuracy    = c(dt_accuracy, nb_accuracy, log_accuracy, knn_accuracy, ann_accuracy),
  Precision   = c(dt_precision, nb_precision, log_precision, knn_precision, ann_precision),
  Recall      = c(dt_recall, nb_recall, log_recall, knn_recall, ann_recall),
  Specificity = c(dt_specificity, nb_specificity, log_specificity, knn_specificity, ann_specificity),
  F1          = c(dt_f1, nb_f1, log_f1, knn_f1, ann_f1),
  AUC         = c(dt_auc, nb_auc, log_auc, NA, ann_auc)
)

kable(results, caption = "Model Performance Comparison", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# 8. Overall Conclusions and Lessons Learned

| Aspect | Key Finding | Real‑world Implication |
|--------|-------------|------------------------|
| **Income prediction (multiple models)** | Logistic Regression and ANN achieved highest AUC (~0.90), while KNN had lower recall. | For imbalanced datasets, AUC and F1 are more informative than raw accuracy. Scaling and encoding are crucial for KNN and ANN. |
| **Clustering (Iris)** | 3 distinct groups, 88% variance explained, silhouette scores ~0.5. | Unsupervised learning can validate natural groupings. Visualisation (PCA, dendrograms) helps interpret clusters. |
| **Outlier detection** | 3.5% consensus outliers, mainly from capital gain and hours per week. | Multiple methods reduce false positives. Outliers may be rare events (e.g., high capital gain) – handle based on analysis goal. |

**Final thoughts for practice:**

- **Never trust a single metric.** Always examine precision, recall, and ROC curve alongside accuracy.
- **Preprocessing matters enormously.** For distance‑based models (KNN, K‑means), scaling is mandatory; for tree‑based models, it is not.
- **Outlier detection is as much an art as a science.** A consensus across diverse methods provides the most reliable flags.
- **Combining different modelling paradigms** (classical statistics, tree‑based, neural networks, distance‑based) gives a well‑rounded understanding of the data.

This analysis demonstrates a complete data science workflow – from data loading and cleaning to modelling, visualisation, and interpretation – using both supervised and unsupervised techniques.

*Report generated on:* `r Sys.Date()`

```{r session}
sessionInfo()
```
```
